---
title: "Analysis of Longitudinal Data"
author: "Shweta Goswami"
date: "8-12-2018"
output: html_document
---

# **Analysis of Longitudinal Data**

## **Analysis of BPRS data**


```{r, include=FALSE}
setwd("~/GitHub/IODS-project/data")
load("~/GitHub/IODS-project/data/BPRSL.Rdata")
library(readr)
BPRS <- read_csv("BPRS.csv")

```

### **Overview of BPRSL (long form) data**

```{r, echo=FALSE}
library(dplyr)
summary(BPRSL)
glimpse(BPRSL)
```

### **Graphical overview of BPRSL data**

```{r, echo=FALSE}
library(tidyr)
library(dplyr)
library(ggplot2)

# Draw the plot
ggplot(BPRSL, aes(x = week, y = bprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(BPRSL$bprs), max(BPRSL$bprs)))
```


### **Graphical overview after standardization**

```{r, echo=FALSE}

# Standardise the variable bprs
BPRSL <- BPRSL %>%
  group_by(week) %>%
  mutate(stdbprs = (bprs - mean(bprs))/sd(bprs) ) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSL)

# Plot again with the standardised bprs
ggplot(BPRSL, aes(x = week, y = stdbprs, linetype = subject)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ treatment, labeller = label_both) +
  scale_y_continuous(name = "standardized bprs")

```


### **Plotting the mean profiles**


```{r, echo=FALSE}


# Number of weeks, baseline (week 0) included
n <- BPRSL$week %>% unique() %>% length()

# Summary data with mean and standard error of bprs by treatment and week 
BPRSS <- BPRSL %>%
  group_by(treatment, week) %>%
  summarise( mean = mean(bprs), se = sd(bprs)/sqrt(n) ) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSS)

# Plot the mean profiles
ggplot(BPRSS, aes(x = week, y = mean, linetype = treatment, shape = treatment)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2)) +
  geom_errorbar(aes(ymin = mean - se, ymax = mean + se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(bprs) +/- se(bprs)")

```


### **Creating boxplot of the new data to check the outlier** 

```{r, echo=FALSE}


# Create a summary data by treatment and subject with mean as the summary variable (ignoring baseline week 0).
BPRSL8S <- BPRSL %>%
  filter(week > 0) %>%
  group_by(treatment, subject) %>%
  summarise( mean=mean(bprs) ) %>%
  ungroup()

# Glimpse the data
glimpse(BPRSL8S)

# Draw a boxplot of the mean versus treatment
ggplot(BPRSL8S, aes(x = treatment, y = mean)) +
  geom_boxplot() +
  stat_summary(fun.y = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(bprs), weeks 1-8")

# Create a new data by filtering the outlier and adjust the ggplot code the draw the plot again with the new data
BPRSL8S1 <- BPRSL8S %>%
  filter(mean < 60)

```



### **Two-sample t-test and Analysis of Variance**

```{r, echo=FALSE}


# Perform a two-sample t-test
t.test(mean ~ treatment, data = BPRSL8S1, var.equal = TRUE)

# Add the baseline from the original data as a new variable to the summary data
BPRSL8S2 <- BPRSL8S %>%
  mutate(baseline = BPRS$week0)

# Fit the linear model with the mean as the response 
fit <- lm(mean ~ baseline + treatment, data = BPRSL8S2)

# Compute the analysis of variance table for the fitted model with anova()
anova(fit)

```

## Analysis of RATS data

```{r, include=FALSE}
load("~/GitHub/IODS-project/data/RATSL.Rdata")
setwd("~/GitHub/IODS-project/data")
library(readr)
RATS <- read_csv("RATS.csv")

```


### **Overview of RATSL (long form) data**

```{r, echo=FALSE}
head(RATSL)
glimpse(RATSL)
```


### **Plotting the RATSL data, identifying the observations in each group**

```{r, echo=FALSE}

# Check the dimensions of the data
dim(RATSL)

# Plot the RATSL data
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 10)) +
  scale_y_continuous(name = "Weight (grams)") +
  theme(legend.position = "top")
```


### **Creating a regression model with Weight as the response variable and Time and Group as explanatory variables**

```{r, echo=FALSE}
# dplyr, tidyr, RATS and RATSL are available

# create a regression model RATS_reg
RATS_reg <- lm(Weight ~ Time + Group, data = RATSL)

# print out a summary of the model
summary(RATS_reg)

```


### **Random intercept model for the same two explanatory variables: Time and Group**

```{r, echo=FALSE}
# dplyr, tidyr, RATS and RATSL are available

# access library lme4
library(lme4)

# Create a random intercept model
RATS_ref <- lmer(Weight ~ Time + Group + (1 | ID), data = RATSL, REML = FALSE)

# Print the summary of the model
summary(RATS_ref)

```

### **Random intercept and random slope model to the rat growth data**

```{r, echo=FALSE}
# dplyr, tidyr, lme4, ggplot2, RATS and RATSL are available

# create a random intercept and random slope model
RATS_ref1 <- lmer(Weight ~ Time + Group + (Time | ID), data = RATSL, REML = FALSE)

# print a summary of the model
summary(RATS_ref1)

# perform an ANOVA test on the two models
anova(RATS_ref1, RATS_ref)

```

### **Random Intercept and Random Slope Model with interaction**

```{r, echo=FALSE}
# dplyr, tidyr, lme4, ggplot2, RATS and RATSL are available

# create a random intercept and random slope model
RATS_ref2 <- lmer(Weight ~ Time * Group + (Time | ID), data = RATSL, REML = FALSE)

# print a summary of the model
summary(RATS_ref2)

# perform an ANOVA test on the two models
anova(RATS_ref2, RATS_ref1)

# draw the plot of RATSL
ggplot(RATSL, aes(x = Time, y = Weight, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Observed weight (grams)") +
  theme(legend.position = "top")

# Create a vector of the fitted values
Fitted <- fitted(RATS_ref2)

# Create a new column fitted to RATSL
RATSL <- RATSL %>%
  mutate(Fitted)

# draw the plot of RATSL
ggplot(RATSL, aes(x = Time, y = Fitted, group = ID)) +
  geom_line(aes(linetype = Group)) +
  scale_x_continuous(name = "Time (days)", breaks = seq(0, 60, 20)) +
  scale_y_continuous(name = "Fitted weight (grams)") +
  theme(legend.position = "top")

```

